Rates and delay times of type Ia supernovae in the Dark Energy Survey

### 2021-03-10 draft.tex

A complementary technique involves measuring the evolution of the volumetric rate of SNe Ia as a function of redshift and comparing this evolution to the average cosmic star-formation history (CSFH) of the Universe \citep{Gal-Yam2004,Strolger2004,Dahlen2004,Dahlen2008, Graur2011,Graur2013, Rodney2014,Frohmaier2019}. This technique (the volumetric rate method) is also applicable to galaxy clusters, for which it is assumed that the star-formation histories are strongly peaked at some past epoch and thus that the rate of SNe Ia in clusters as a function of redshift is a more direct measure of the DTD \citep{Maoz2004,Maoz2010,Friedman2018,Freundlich2021}. These studies have, almost ubiquitously, found $\beta$ to be consistent with $-1$.

@@

The majority of observational evidence based on studies of SN Ia populations thus points towards a DD scenario for most, if not all, SNe Ia. However, finding self-consistent explosion models that recreate the observed luminosity function as well as correlations between luminosity, light curve parameters, and host galaxy properties has proven difficult. In particular, simulations based around explosions of $M_{\mathrm{Ch}}$ WDs fail to reproduce the observed relationship between SN luminosity and light curve duration as well as "peculiar" objects \citep{Sim2013,Blondin2017}. In recent years, attention directed towards explosions of sub-$M_{\mathrm{Ch}}$ WDs triggered by double detonations has led to promising results \citep[e.g.][]{Shen2017,Shen2018,Townsley2019,Gronow2020,Shen2021} although they still struggle to match observations at late times in the light curve evolution \citep{Shen2021,Gronow2021}. An additional factor in support of the double-detonation model is that the SN luminosity is related to the mass of the primary WD which can be related to its age, thereby providing an explanation for observed relation between light curve stretch and stellar age \citep{Rigault2013,Rigault2018,Nicolas2020}.

@@

\begin{table}
	\centering
	\caption{Numbers of field galaxies and SN hosts passing various quality cuts.}
	\label{tab:cuts}
	\begin{tabular}{lccr} % four columns, alignment for each
		\hline
		Cut & N (field galaxies)  & N (SN hosts)\\
		\hline
		Chosen fields & 1364311  & 1766\\
	    Kron mag in all bands & 1069004  & 1764 \\
	    Masked chips & 816950  & - \\
	    Has ugrizJHK photometry & 545748  & -\\
	    Edge of chip & 481731 & 1716 \\
	    Star/galaxy separation & 400051 &  1678\\
	    Has good redshift & 395034 & 1678 \\
	    SNR $\geq 3$& 338256  & 1678 \\
	    $m_r \leq 24.5$ & 196109 &  1678 \\
	    $0.2 \leq z \leq 0.6$ & 48177 & 1414\\

@@

\label{sec:data}
\subsection{Dark Energy Survey supernova programme \label{subsec:des}}
The Dark Energy Survey Supernova Programme (DES-SN) is a transient survey based on five six-month seasons of observations with the Dark Energy Camera (DECam; \citealt{Flaugher2015}), designed primarily to measure the light curves of SNe Ia for use as cosmological distance indicators. Transients were detected and processed using a difference imaging program \citep{Kessler2015} and vetted using a machine learning (ML) algorithm \citep{Goldstein2015}. Spectral follow-up of live SN candidates was performed on a suite of large optical telescopes \citep{Smith2020a}, leading to the initial publication of a measurement of cosmological parameters using 207 spectroscopically confirmed SNe Ia (\citealt{DESCollaboration2018a}, and references therein). 
\subsection{Supernovae \label{subsec:host_sample}}
\subsubsection{Photometric classification and quality cuts \label{subsubsec:sn_classify}}
With over 30,000 transients passing ML selection it was not possible to obtain spectral follow up of every object. Instead we make use of photometric classification in the form of the recurrent neural network classifier \texttt{superNNova} (SNN; \citealt{Moller2019}), mirroring the approach of \citet{Scolnic2020}. We train SNN using a large suite of simulated multi-band SN light curves. \textbf{Waiting on final sample and a few words from Maria/Anais to complete this bit.} SNe Ia are simulated based on the \texttt{SALT2} model \citep{Guy2007} trained on the Joint Lightcurve Analysis (JLA) data set \citep{Betoule2014}. CCSNe are generated from the templates of \citet{Vincenzi2019}. We select SNe with a threshold probability of $P(\mathrm{Ia})\geq0.5$, although our analysis is not sensitive to this choice, since the vast majority of SNe receive classifications close to 1 or 0. The photometrically classified SNe Ia are then passed through the \texttt{SALT2} framework light curve fitter and are subject to quality cuts on the light curve width (stretch; $x_1$) and colour ($c$) parameters in an identical manner to \citet{Vincenzi2020}: $-3 \leq x_1 \leq 3$, $-0.3 \leq c \leq 0.3$. This selection helps to reduce the potential contamination from core-collapse SNe (CCSNe) and outlying thermonuclear SNe. We impose further cuts on the quality of those measured parameters, which are: the uncertainty on $x_1$: $\sigma_{x_1} <1$; the uncertainty on the date at which the light curve peaked: $\sigma_{t_{\mathrm{peak}}} <2$; the SALT2 fit probability $>0.001$.

\subsubsection{Host galaxy selection \label{subsubsec:sn_hosts}}
Host galaxies for all transients in DES-SN are retrieved from the DES-SN Deep catalogue, a galaxy catalogue from the $g, r, i, z$-band coadded images of the ten DES-SN fields which was presented in \citet{Wiseman2020}. SNe are associated to galaxies using the directional light radius (DLR) method \citep[e.g.][]{Sullivan2006,Gupta2016}. As is standard across DES-SN analyses, we require a SN-host separation to DLR ratio ($d_{\mathrm{DLR}}$) less than 4 in order to classify a galaxy as a host. In the case where multiple galaxies lie within $d_{\mathrm{DLR}}<4$, the object with the lower or lowest value is taken to be the host. Spectroscopic redshifts for host galaxies come from the OzDES Global Redshift Catalog (GRC; \footnote{\url{https://docs.datacentral.org.au/ozdes/overview/dr2/}} \citealt{Yuan2015,Childress2017,Lidman2020}), which comprises galaxies targetted as DES-SN hosts with the OzDES programme as well as redshifts from legacy catalogues in the DES-SN fields.

@@

The posterior distributions for the fit parameters are found in Figs. \ref{fig:corner_beta_norm} and \ref{fig:corner_beta_norm_tp} for the fixed and free $t_{\mathrm{p}}$ fits respectively, and the posterior means and standard deviations are summarised in Table \ref{tab:dtd_results}. In Fig. \ref{fig:rate_fitted} we show the SN rate per galaxy predicted by the model assuming the best fitting DTD parameters from Table \ref{tab:dtd_results} compared to the data as presented in Section \ref{sec:rates}. The model provides a good fit across a wide range of stellar mass. The model clearly diverges from the simple log-linear fit, describing better the suppressed rate at the low and high ends of the stellar mass range, as well as the enhanced rate around $10^{10}$ M$_{\odot}$. This improvement provides evidence to support the DTD*SFH model. However, the model prediction still diverges from the data at high stellar masses ($>10^{11.25}$ M$_{\odot}$), suggesting that these data cannot be explained by our model, but are driven by further processes that we have not included. We discuss these in Section \ref{subsec:discussion}.

@@

An alternative view on how the SN Ia rate relates to environment can be gained by calculating the rate of SNe per year per unit stellar mass: SNuM. A measurement of the SNuM is achieved by summing the number of SNe in each stellar mass bin, and dividing by the total stellar mass of field galaxies in that bin. We show the SNuM as a function of galaxy stellar mass in Fig.~ \ref{fig:snum}, along with the measurements from SNe discovered in SDSS spectra \citep{Graur2013} at $z\sim0.1$. We also calculate the SNuM predicuted by our model for each simulated galaxy at a range of redshifts and plot these as curves in Fig.~\ref{fig:snum}. Our model accurately recovers the SNuM measured in DES for the redshift and stellar mass range of the data. While the data from \citet{Graur2013} lies slightly higher than our model prediction, the difference is consistent with differences in assumed SFHs and stellar population templates as well as SN detection method.
\section{The delay-time distribution as a function of supernova properties}
\label{sec:split_x1_c}

@@

The SN Ia rates generated by the posterior predictive checks are shown in Fig. \ref{fig:rate_fitted_split_x1_2c_DTD}. The best fitting values are somewhat consistent with one another, with $t_1 = 0.65 \pm 0.1$~Gyr and $t_2 = 0.85 \pm 0.15$~Gyr, corresponding to a relatively sharp transition between the two populations. Our $t_1$ and $t_2$ are consistent with the transition found by \citet{Brandt2010} who found high-stretch SNe to be confined to delay times $\lesssim 0.4$~Gyr while high-stretch SNe displayed delay times $\gtrsim 2.4$~Gyr. These results are somewhat surprising, as we might expect that the transition is not so sharp but a gradual correlation of delay time with the average stretch of SNe caused by the age vs WD-mass distribution, as predicted by \citet{Shen2021}. However we note that the intrinsic stretch distribution may not be a simple Gaussian, but could instead be bimodal \citep{Scolnic2016,Popovic2021}. The correlation of progenitor age and WD mass does not easily reproduce such a bimodal distribution, which may suggest that the sub-$M_{\mathrm{Ch}}$ channel cannot account for all SNe Ia. We also note that our method models the average SFH of galaxies and that the DTD is a statistical distribution: our results do not rule out high-stretch SNe arising from old progenitors, but do suggest that such a scenario is unlikely.

Our transition ages $t_1$ and $t_2$

@@

One possibility is that sub-luminous SNe are more numerous in such predominantly old stellar populations, such that the fraction passing our light curve cuts is lower than the rest of the sample. Such a phenomenon could be caused by SN Ia sub-classes such as SN 1991bg-like SNe, which are known to explode exclusively in old environments \citep{Perets2010} at delay times as long as $>6$~Gyr \citep{Panther2019}. If the SN 1991bg-like objects are drawn from the same parent population of WDs as the "normal" SNe Ia, they could explain the lack thereof in the sample. We check this briefly by re-examining the objects left out of our sample due to \texttt{SALT2} cuts. We find that 75\% of objects left out lie in hosts with $M_*>10^{10}$ M$_{\odot}$, while this percentage is 65\% for the objects that pass the cuts. However, of the objects in high-mass galaxies that fail the cuts, only 11 (10\%) fail due to their $x_1$ lying below $-3$, and a total of 20 have $x_1 <-2$ which is the regime populated by SN 1991bg-like objects which is well short of the factor of two by which the DTD model over-predicts the data in the highest mass bin.

>>>>

A complementary technique involves measuring the evolution of the volumetric rate of SNe Ia as a function of redshift and comparing this evolution to the average cosmic star-formation history (CSFH) of the Universe \citep{Gal-Yam2004,Strolger2004,Dahlen2004,Dahlen2008, Graur2011,Graur2013, Rodney2014,Frohmaier2019}. This technique (the volumetric rate method) is also applicable to galaxy clusters, for which it is assumed that the star-formation histories are strongly peaked at some past epoch and thus that the rate of SNe Ia in clusters as a function of redshift is a more direct measure of the DTD \citep{Maoz2004,Maoz2010,Friedmann2018,Freundlich2021}. These studies have, almost ubiquitously, found $\beta$ to be consistent with $-1$.

@@

The majority of observational evidence based on studies of SN Ia populations thus points towards a DD scenario for most, if not all, SNe Ia. However, finding self-consistent explosion models that recreate the observed luminosity function as well as correlations between luminosity, light curve parameters, and host galaxy properties has proven difficult. In particular, simulations based around explosions of $M_{\mathrm{Ch}}$ WDs fail to reproduce the observed relationship between SN luminosity and light curve duration as well as "peculiar" objects \citep{Sim2013,Blondin2017}. In recent years, attention directed towards explosions of sub-$M_{\mathrm{Ch}}$ WDs triggered by double detonations has led to promising results \citep[e.g.][]{Shen2017,Shen2018,Townsley2019,Gronow2020,Shen2021} although they still struggle to match observations at late times in the light curve evolution \citep{Shen2021,Gronow2021}. An additional factor in support of the double-detonation model is that the SN luminosity is related to the mass of the primary WD, which itself is likely to be related to its age (although this relation is probably complicated by other factors such as accretion rate, metallicity, and the composition of the companion), thereby providing an explanation for observed relation between light curve stretch and stellar age \citep{Rigault2013,Rigault2018,Nicolas2020}.

@@

\begin{table}
	\centering
	\caption{Numbers of field galaxies and SN hosts passing various quality cuts. SNe are derived from must already have passed the masked chips and good redshift cuts before reaching this stage.}
	\label{tab:cuts}
	\begin{tabular}{lccr} % four columns, alignment for each
		\hline
		Cut & N (field galaxies)  & N (SN hosts)\\
		\hline
		Chosen fields & 1364311  & 1441\\
	    Kron mag in all bands & 1069004  & 1439 \\
	    Masked chips & 816950  & - \\
	    Has ugrizJHK photometry & 545748  & -\\
	    Edge of chip & 481731 & 1401 \\
	    Star/galaxy separation & 400051 &  1259\\
	    Has good redshift & 395034 & - \\
	    SNR $\geq 3$& 338256  & 1259 \\
	    $m_r \leq 24.5$ & 196109 &  1254 \\
	    $0.2 \leq z \leq 0.6$ & 48177 & 881\\

@@

\label{sec:data}
\subsection{Dark Energy Survey supernova programme \label{subsec:des}}
The Dark Energy Survey Supernova Programme (DES-SN) is a transient survey based on five six-month seasons of observations with the Dark Energy Camera (DECam; \citealt{Flaugher2015}), designed primarily to measure the light curves of SNe Ia for use as cosmological distance indicators. Transients were detected and processed using a difference imaging program \citep{Kessler2015} and vetted using a machine learning (ML) algorithm \citep{Goldstein2015}. Spectral follow-up of live SN candidates was performed on a suite of large optical telescopes \citep{Smith2020a}, leading to the initial publication of a measurement of cosmological parameters using 207 spectroscopically confirmed SNe Ia (\citealt{DESCollaboration2018a}, and references therein). Spectroscopic redshifts for host galaxies come from the OzDES Global Redshift Catalog (GRC; \footnote{\url{https://docs.datacentral.org.au/ozdes/overview/dr2/}} \citealt{Yuan2015,Childress2017,Lidman2020}), which comprises galaxies targetted as DES-SN hosts with the OzDES programme as well as redshifts from legacy catalogues in the DES-SN fields.
\subsection{Supernovae \label{subsec:host_sample}}
\subsubsection{Photometric classification and quality cuts \label{subsubsec:sn_classify}}
With over 30,000 transients passing ML selection it was not possible to obtain spectral follow up of every object. Instead we make use of photometric classification in the form of the recurrent neural network classifier \texttt{superNNova} (SNN; \citealt{Moller2019}), mirroring the approach of \citet{Scolnic2020}. We train SNN using a large suite of simulated multi-band SN light curves. \textbf{Waiting on final sample and a few words from Maria/Anais to complete this bit.} SNe Ia are simulated based on the \texttt{SALT2} model \citep{Guy2007} trained on the Joint Lightcurve Analysis (JLA) data set \citep{Betoule2014}. Synthetic CCSNe light curves are generated from the templates of \citet{Vincenzi2019}. Given the training set of synthetic light curves, SNN is able to return a classification for observed light curves. Before classifying, we remove transients with variability in multiple seasons (likely active galactic nuclei or superluminous SNe). SNN is then run on every transient with a host galaxy redshift (see Section \ref{subsubsec:sn_hosts}). We define SNe with a threshold probability of $P(\mathrm{Ia})\geq0.5$ as photometrically classified SNe Ia, although our analysis is not sensitive to this choice, since the vast majority of SNe receive classifications close to 1 or 0 \citep{Moller2019}. The photometrically classified SNe Ia are then passed through the \texttt{SALT2} framework light curve fitter and are subject to quality cuts on the light curve width (stretch; $x_1$) and colour ($c$) parameters in an identical manner to \citet{Vincenzi2020}: $-3 \leq x_1 \leq 3$, $-0.3 \leq c \leq 0.3$. This selection helps to reduce the potential contamination from core-collapse SNe (CCSNe) and outlying thermonuclear SNe. We impose further cuts on the quality of those measured parameters, which are: the uncertainty on $x_1$: $\sigma_{x_1} <1$; the uncertainty on the date at which the light curve peaked: $\sigma_{t_{\mathrm{peak}}} <2$; the SALT2 fit probability $>0.001$. The number of SNe passing each stage of these cuts is displayed in Table 2 of \citet{Vincenzi2020} and reduces the $\sim30,000$ objects to 1604 SNe. With the addition of the photometric classifier we are left with 1441 SNe with which we proceed to inspect their host galaxies.

\subsubsection{Host galaxy selection \label{subsubsec:sn_hosts}}
Host galaxies for all transients in DES-SN are retrieved from the DES-SN Deep catalogue, a galaxy catalogue from the $g, r, i, z$-band coadded images of the ten DES-SN fields which was presented in \citet{Wiseman2020}. SNe are associated to galaxies using the directional light radius (DLR) method \citep[e.g.][]{Sullivan2006,Gupta2016}. As is standard across DES-SN analyses, we require a SN-host separation to DLR ratio ($d_{\mathrm{DLR}}$) less than 4 in order to classify a galaxy as a host. In the case where multiple galaxies lie within $d_{\mathrm{DLR}}<4$, the object with the lower or lowest value is taken to be the host.

@@

The joint and marginal posterior distributions for the fit parameters are found in Figs. \ref{fig:corner_beta_norm} and \ref{fig:corner_beta_norm_tp} for the fixed and free $t_{\mathrm{p}}$ fits respectively, and the posterior means and standard deviations are summarised in Table \ref{tab:dtd_results}.  Fig. \ref{fig:corner_beta_norm} shows that there is a correlation between $\beta$ and $A$ which is not seen in the fit with $t_{\mathrm{p}}$ as a free parameter. This is likely caused by $\beta$ instead correlating with $t_{\mathrm{p}}$, and being more poorly constrained in the three-parameter fit. In Fig. \ref{fig:rate_fitted} we show the SN rate per galaxy predicted by the model assuming the best fitting DTD parameters from Table \ref{tab:dtd_results} compared to the data as presented in Section \ref{sec:rates}. The model provides a good fit across a wide range of stellar mass. The model clearly diverges from the simple log-linear fit, describing better the enhanced rate around $10^{10}$ M$_{\odot}$ and surpression at high masses. This improvement provides evidence to support the DTD*SFH model. However, the model prediction still diverges from the data at high stellar masses ($>10^{11.25}$ M$_{\odot}$), suggesting that these data cannot be explained by our model, but are driven by further processes that we have not included. We discuss these in Section \ref{subsec:discussion}.

@@

An alternative view on how the SN Ia rate relates to environment can be gained by calculating the rate of SNe per year per unit stellar mass: SNuM. A measurement of the SNuM is achieved by summing the number of SNe in each stellar mass bin, and dividing by the total stellar mass of field galaxies in that bin. We show the SNuM as a function of galaxy stellar mass in Fig.~ \ref{fig:snum}, along with the measurements from SNe discovered in SDSS spectra \citep{Graur2013} at $z\sim0.1$ and at $z\sim0$ in LOSS as presented in \citet{Li2011a}. We also calculate the SNuM predicuted by our model for each simulated galaxy at a range of redshifts and plot these as curves in Fig.~\ref{fig:snum}. Our model accurately recovers the SNuM measured in DES for the redshift and stellar mass range of the data. While the data from LOSS and SDSS lies slightly higher than our model prediction, the difference is consistent with differences in assumed SFHs, IMFs and stellar population templates as well as SN detection method. Any true difference between predicted and observed SNuM could be caused by an evolution of the SN Ia production efficiency as a function of redshift; we defer such an investigation to future work with more refined models and measurements.
\section{The delay-time distribution as a function of supernova properties}
\label{sec:split_x1_c}

@@

The SN Ia rates generated by the posterior predictive checks are shown in Fig. \ref{fig:rate_fitted_split_x1_2c_DTD}. The best fitting values are somewhat consistent with one another, with $t_1 = 0.65 \pm 0.1$~Gyr and $t_2 = 0.85 \pm 0.15$~Gyr, corresponding to a relatively sharp transition between the two populations. Our $t_1$ and $t_2$ are consistent with the transition found by \citet{Brandt2010} who found high-stretch SNe to be confined to delay times $\lesssim 0.4$~Gyr while high-stretch SNe displayed delay times $\gtrsim 2.4$~Gyr. On the other hand, the transition found here appears to occur at much shorter delay times than the several Gyrs found by \citet{Rose2019}. These results are somewhat surprising, as we might expect that the transition is not so sharp but a gradual correlation of delay time with the average stretch of SNe caused by the age vs WD-mass distribution, as predicted by \citet{Shen2021}. However we note that the intrinsic stretch distribution may not be a simple Gaussian, but could instead be bimodal \citep{Scolnic2016,Popovic2021}. The correlation of progenitor age and WD mass does not easily reproduce such a bimodal distribution, which may suggest that the sub-$M_{\mathrm{Ch}}$ channel cannot account for all SNe Ia. We also note that our method models the average SFH of galaxies and that the DTD is a statistical distribution: our results do not rule out high-stretch SNe arising from old progenitors, but do suggest that such a scenario is unlikely.

@@

One possibility is that sub-luminous SNe are more numerous in such predominantly old stellar populations, such that the fraction passing our light curve cuts is lower than the rest of the sample. Such a phenomenon could be caused by SN Ia sub-classes such as SN 1991bg-like SNe, which are known to explode exclusively in old environments \citep{Perets2010} at delay times as long as $>6$~Gyr \citep{Panther2019}, and are typically sub-luminous with fast decline rates \citep{Gonzalez-Gaitan2014}. If the SN 1991bg-like objects are drawn from the same parent population of WDs as the "normal" SNe Ia, they could explain the lack thereof in the sample. We check this briefly by re-examining the objects left out of our sample due to \texttt{SALT2} cuts. We find that 75\% of objects left out lie in hosts with $M_*>10^{10}$ M$_{\odot}$, while this percentage is 65\% for the objects that pass the cuts. However, of the objects in high-mass galaxies that fail the cuts, only 11 (10\%) fail due to their $x_1$ lying below $-3$, and a total of 20 have $x_1 <-2$ which is the regime populated by SN 1991bg-like objects which is well short of the factor of two by which the DTD model over-predicts the data in the highest mass bin.
